---
title: "Chapter 21: Competing Risks Models and Chapter 22: Survival Trees"
output: html_notebook
---

## Libraries

```{r setup}
library(ggplot2)
library(scales)
library(dplyr)
library(survival)
library(survminer)
library(lubridate)
library(tidyr)
library(stringr)
library(wesanderson)
library(randomForest)
library(xtable)
library(rpart)
library(rpart.plot)
library(glmnet)
```

## Simulated Survival Data (ESL Dataset)

```{r}
load("../data/mixture.example.RData")
set.seed(100)
df <- data.frame(x1 = mixture.example$x[,1], 
                 x2 = mixture.example$x[,2]) %>%
  mutate(r1 = 2 - 1.5 * x1 + 0.2 * x2,    # risk score for outcome 1 (discharge)
         r2 = -1 + 0.8 * x1 + 0.0 * x2,    # risk score for outcome 2 (death)
         y1 = rexp(length(x1), exp(r1)),  # outcome 1 (discharge)
         y2 = rexp(length(x1), exp(r2)),  # outcome 2 (death)
         yC = rexp(length(x1), 1)) %>%  # censoring distribution (transfer)
  mutate(ftime1C = if_else(y1 < yC, y1, yC),  # follow-up time for outcome 1, uninformative censoring
         ctime1C = if_else(y1 < yC, 1, 0),    
         ftime2C = if_else(y2 < yC, y2, yC),  # follow-up time for outcome 2, uninformative censoring
         ctime2C = if_else(y2 < yC, 1, 0),    
         ftime12 = case_when(y1 < y2 & y1 < yC ~ y1, 
                             y2 < yC ~ y2, 
                             TRUE ~ yC), 
         ctime12 = case_when(y1 < y2 & y1 < yC ~ "discharge", 
                             y2 < yC ~ "death", 
                             TRUE ~ "censored"))
cor(df$y1, df$y2)
summary(as.factor(df$ctime12))
```

Plot of raw disease severity and socioeconomic scores.

```{r}
p <- ggplot(df) + geom_point(aes(x = x1, y = x2), colour = "gray40") + 
  xlab("Disease Severity Score (x1)") + 
  ylab("Social Determinants Score (x2)") + 
  theme_bw()
print(p)
ggsave(p, filename = "../img/esl-just-data-no-classes.png", height=4, width=4, units="in", dpi=300)
```

Plot survival times for a few different patients.

```{r}
t <- seq(0, 5, 0.01)
y1_60 <- dexp(t, exp(df$r1[60]))
y1_32 <- dexp(t, exp(df$r1[32]))
y2_60 <- dexp(t, exp(df$r2[60]))
y2_32 <- dexp(t, exp(df$r2[32]))
yC <- dexp(t, 1)
df_y60 <- data.frame(t = rep(t, 3), ft = c(y1_60, y2_60, yC), 
                     group = c(rep("discharge", length(y1_60)), 
                               rep("death", length(y2_60)), 
                               rep("censoring",length(yC))))
df_y32 <- data.frame(t = rep(t, 3), ft = c(y1_32, y2_32, yC), 
                     group = c(rep("discharge", length(y1_32)), 
                               rep("death", length(y2_32)), 
                               rep("censoring",length(yC))))

p1 <- ggplot(df_y60) + geom_line(aes(x = t, y = ft, colour = group, linetype = group)) + 
  theme_bw() + ylab("f(t) for Person 60") + xlab("t") + 
  scale_colour_manual(name = "Outcome", breaks = c("death", "discharge", "censoring"), 
                      values = c("black", "black", "gray70")) + 
  scale_linetype_manual(name = "Outcome", breaks = c("death", "discharge", "censoring"), 
                        values = c("dashed", "solid", "solid"))

p2 <- ggplot(df_y32) + geom_line(aes(x = t, y = ft, colour = group, linetype = group)) + 
  theme_bw() + ylab("f(t) for Person 32") + xlab("t") + 
  scale_colour_manual(name = "Outcome", breaks = c("death", "discharge", "censoring"), 
                      values = c("black", "black", "gray70")) + 
  scale_linetype_manual(name = "Outcome", breaks = c("death", "discharge", "censoring"), 
                        values = c("dashed", "solid", "solid"))

print(p1)
print(p2)

ggsave(p1, filename = "../img/cmprsk-example-person-60.png", height = 3, width = 5, units = "in", dpi = 300)
ggsave(p2, filename = "../img/cmprsk-example-person-32.png", height = 3, width = 5, units = "in", dpi = 300)
```

Sample outcomes for person 32.

```{r}
n = 100
y1_32 <- rexp(n, exp(df$r1[32]))
y2_32 <- rexp(n, exp(df$r2[32]))
yC_32 <- rexp(n, 1)
df_samples <- data.frame(y = c(y1_32, y2_32, yC_32), 
                         n = c(1:n, 1:n, 1:n),
                         group = c(rep("discharge", n), 
                                   rep("death", n), 
                                   rep("censoring", n)))
df_samples <- df_samples %>%
  arrange(y) %>%
  group_by(n) %>%
  slice(1)

p <- ggplot(df_samples) + 
  geom_histogram(aes(x = y, fill = group, colour = group), position = "stack", alpha = 0.8, bins = 30) + 
  theme_bw() + xlab("Follow-up Time") + ylab("Count") + 
  scale_fill_manual(name = "Outcome", breaks = c("death", "discharge", "censoring"), 
                      values = c("black", "white", "gray70")) + 
  scale_colour_manual(name = "Outcome", breaks = c("death", "discharge", "censoring"), 
                      values = c("black", "black", "gray30"))
print(p)
ggsave(p, filename = "../img/cmprsk-example-person-32-samples.png", height = 3, width = 5, 
       units = "in", dpi = 300)
```

What our real dataset would look like.

```{r}
df_real <- df %>% select(x1, x2, ftime12, ctime12)
df_real[10:20,]
```

Simulation 1: death and censoring only

```{r}
m2C <- coxph(Surv(ftime2C, ctime2C) ~ x1 + x2, data = df)
m2C
summary(as.factor(df$ctime2C))

s <- survfit(Surv(ftime2C, ctime2C) ~ 1, data = df)
p <- ggsurvplot(s)

png(filename = "../img/cmprsk-example-km-sim-1.png", height=4, width=5, units="in", res=300)
print(p)
dev.off()

print(p)
```

Simulation 2: death treating discharge same as censoring

```{r}
m12b <- coxph(Surv(ftime12, ctime12 == "death") ~ x1 + x2, data = df)
m12b
summary(as.factor(df$ctime12 == "death"))

s <- survfit(Surv(ftime12, ctime12 == "death") ~ 1, data = df)
p <- ggsurvplot(s)

png(filename = "../img/cmprsk-example-km-sim-2.png", height=4, width=5, units="in", res=300)
print(p)
dev.off()

print(p)
```

Kaplan-Meier curves (overall). 

```{r}
df_km <- data.frame(ftime = c(df$ftime2C, df$ftime12), 
                    ctime = c(df$ctime2C, ifelse(df$ctime12 == "death", 1, 0)), 
                    group = c(rep("no discharges", dim(df)[1]), rep("discharges\nas censoring", dim(df)[1])), 
                    x1 = c(df$x1, df$x1), x2 = c(df$x2, df$x2))

s <- survfit(Surv(ftime, ctime) ~ group, data = df_km, stype = 1, ctype = 1)
s_df <- data.frame(time = s$time, surv = s$surv, cumhaz = s$cumhaz, 
                   censor = s$n.censor, group = rep(names(s$strata), each = 200))
p <- ggplot(s_df) + 
  geom_step(aes(x = time, y = surv, colour = group), size = 1) + 
  theme_light() + xlab("Time") + ylab("Survival Probability (%)") + ylim(0, 1) + 
  geom_point(aes(x = time, y = surv, colour = group), data = subset(s_df, censor == 1), shape = 3, size = 3)
ggsave(p, filename = "../img/cmprsk-example-km-all.png", height=4, width=6, units="in", dpi=300)
print(p)

p <- ggplot(s_df) + 
  geom_step(aes(x = time, y = cumhaz, colour = group), size = 1) + 
  theme_light() + xlab("Time") + ylab("Cumulative Hazard") +
  geom_point(aes(x = time, y = cumhaz, colour = group), data = subset(s_df, censor == 1), shape = 3, size = 3)
ggsave(p, filename = "../img/cmprsk-example-na-all.png", height=4, width=6, units="in", dpi=300)
print(p)
```

Kaplan-Meier curves for discharge and death.

```{r}
s <- survfit(Surv(ftime12, ctime12 == "death") ~ 1, data = df, stype = 1, ctype = 1)
s_df <- data.frame(time = s$time, surv = s$surv, cumhaz = s$cumhaz, 
                   censor = s$n.censor, model = "death")
s2 <- survfit(Surv(ftime12, ctime12 == "discharge") ~ 1, data = df, stype = 1, ctype = 1)
s_df2 <- data.frame(time = s2$time, surv = s2$surv, cumhaz = s2$cumhaz, 
                   censor = s2$n.censor, model = "discharge")
s_df <- rbind(s_df, s_df2)

p <- ggplot(s_df) + 
  geom_step(aes(x = time, y = 1 - surv, colour = model), size = 1) + 
  theme_light() + xlab("Time") + ylab("Probability (Kaplan-Meier)") + ylim(0, 1) + 
  # geom_point(aes(x = time, y = surv, colour = model), 
  # subset = subset(s_df, censor == 1), shape = 3, size = 3) + 
  scale_colour_discrete(name = "Outcome")
ggsave(p, filename = "../img/cmprsk-example-km-two-outcomes.png", height=4, width=6, units="in", dpi=300)
print(p)

s3 <- cuminc(df$ftime12, as.factor(df$ctime12), cencode = "censored")
s3_df <- data.frame(time = c(s3$`1 death`$time, s3$`1 discharge`$time), 
                    est = c(s3$`1 death`$est, s3$`1 discharge`$est), 
                    model = c(rep("death", length(s3$`1 death`$time)), 
                              rep("discharge", length(s3$`1 discharge`$time))))
p <- ggplot(s3_df) + 
  geom_step(aes(x = time, y = est, colour = model), size = 1) + 
  theme_light() + xlab("Time") + ylab("Cumulative Incidence") + ylim(0, 1) + 
  scale_colour_discrete(name = "Outcome")
ggsave(p, filename = "../img/cmprsk-example-cuminc-two-outcomes.png", height=4, width=6, units="in", dpi=300)
print(p)
```

Fix factor issue.

```{r}
df$ctime12 = as.factor(df$ctime12)
```

Cox model for simulated data.

```{r}
m_cox <- coxph(Surv(ftime12, ctime12) ~ x1 + x2, data = df, id = row.names(df))
summary(m_cox)
```

Fine-Gray model for simulated data. 

```{r}
m_fg <- crr(df$ftime12, df$ctime12, select(df, c("x1", "x2")), failcode = "death", cencode = "censored")
summary(m_fg)
m_fg2 <- crr(df$ftime12, df$ctime12, select(df, c("x1", "x2")), failcode = "discharge", cencode = "censored")
summary(m_fg2)
```

Predictions.

```{r}
p_cox <- predict(m_cox, type = "risk")
```

## Chapter 22: Survival Trees

Building a single survival tree from our simulated data.

```{r}
palette <- colorRampPalette(colors=c("#00BFC4", "#F8766D"))
tree_m <- rpart(Surv(ftime12, ctime12 == "death") ~ x1 + x2, data = df)
rpart.plot(tree_m, box.palette=palette(20))
```

```{r}
library(partykit)
tree_m2 = as.party(tree_m)
png("../img/survival-tree-example-simulated.png", height=9, width=15, units="in", res=400)
plot(tree_m2, )
dev.off()
```

Competing risks analysis.

```{r}
df2 <- df
df2$ctime12 = as.factor(ifelse(df2$ctime12 == "censored", 0, df2$ctime12))
wihs.obj <- rfsrc(Surv(ftime12, ctime12) ~ x1 + x2, data=df, nsplit=4, ntree=100)
#  Plot the ensemble cs-CIF, and cs-CHF.
plot.competing.risk(wihs.obj)
```

